Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-06-19 pirouette_article.tex
Process feedback

For example, one can add novel diversification models 
by writing a BEAST2 package that contains the likelihood 
formula of a phylogeny under the novel diversification model, 
i.e. the prior probability of a species tree.
Many diversification models (and their associated probability algorithms) 
have been developed, e.g., models in which diversification is time-dependent [\cite{nee1994reconstructed,rabosky2008explosive}],
or diversity-dependent [\cite{etienne2012diversity}],
or where diversification rates change for specific lineages 
and their descendants [\cite{etienne2012conceptual, rabosky2014automatic, alfaro2009nine}].
Other models treat speciation as a process that takes 
time [\cite{rosindell2010protracted, etienne2012prolonging, lambert2015reconstructed}],
or where diversification rate
depends on one or more traits [\cite{maddison2007estimating, fitzjohn2012diversitree}].
Despite such a rich theoretical landscape, only a few of these 
diversification models are available as tree priors in BEAST2
% 00005
\new{
  Despite the fact that some models are already included as tree priors 
  in BEAST2 (e.g. coalescent models, BD skyline, etc.) 
  many others for which a likelihood formula is already available 
  in literature, are not implemented yet.
}
\new{
  BEAST2 allows to extend its functionality by plug-ins, such
  as [RJCB: name some plug-ins with refs to literature].
}

When a novel diversification model is introduced,
its performance in inference should be tested.
Part of a model's performance is its ability to 
recover parameters from simulated data with known 
parameters (e.g. [\cite{etienne2014estimating}]), 
where ideally the estimated parameter values closely match the known/true values.
Even when a diversification model passes this procedure, 
it is not necessarily used as tree prior in Bayesian inference.
Bayesian phylogenetic inference often requires 
that the prior probability of the phylogeny 
according to the diversification model has to be computed millions of times. 
Therefore, biologically interesting but computationally expensive tree priors 
are often not implemented, and simpler priors are used instead. 
This is not necessarily problematic, when the data are very informative 
\new{or when the prior is truly uninformative}, 
as this will reduce the influence of the tree prior.
However, the assumption that tree prior choice is of low \new{impact} 
must first be verified.

@@

\begin{lstlisting}[language=R]
library(pirouette)

# Create phylogeny
set.seed(314)
phylogeny <- create_exemplary_dd_tree(
  n_taxa = 6, 
  crown_age = 10,
  extinction_rate = 0.1
)

# Setup pirouette
pir_params <- create_std_pir_params(
  folder_name = "example_30"
)

@@

# Save
library(ggplot2)
pir_plot(pir_out) +
  ggtitle("pirouette example 30") +
  ggsave("errors.png", width = 7, height = 7)

pir_save(
  phylogeny = phylogeny,
  pir_params = pir_params,
  pir_out = pir_out,
  folder_name = "example_30"
)
\end{lstlisting}

>>>>

For example, one can add novel diversification models 
by writing a BEAST2 package that contains the likelihood 
formula of a phylogeny under the novel diversification model, 
i.e. the prior probability of a species tree.
Many diversification models (and their associated probability algorithms) 
have been developed, e.g., models in which diversification is time-dependent [\cite{nee1994reconstructed,rabosky2008explosive}],
or diversity-dependent [\cite{etienne2012diversity}],
or where diversification rates change for specific lineages 
and their descendants [\cite{etienne2012conceptual, rabosky2014automatic, alfaro2009nine}].
Other models treat speciation as a process that takes 
time [\cite{rosindell2010protracted, etienne2012prolonging, lambert2015reconstructed}],
or where diversification rate
depends on one or more traits [\cite{maddison2007estimating, fitzjohn2012diversitree}].
Despite such a rich theoretical landscape, only a few of these 
diversification models are available as tree priors in BEAST2
% 00005
\new{
  Despite the fact that some models are already included as tree priors 
  in BEAST2 (e.g. birth-death models, coalescent models, coalescent skyline, etc.) 
  many others for which a likelihood formula is already available 
  in literature, are not implemented yet.
}
\new{
  BEAST2 allows to extend its functionality by plug-ins, such
  as BD skyline (ref) [RJCB: name some other plug-ins with 
  refs to literature].
}

When a novel diversification model is introduced,
its performance in inference should be tested.
Part of a model's performance is its ability to 
recover parameters from simulated data with known 
parameters (e.g. [\cite{etienne2014estimating}]), 
where ideally the estimated parameter values closely match the known/true values.
Even when a diversification model passes this procedure, 
it is not necessarily used as tree prior in Bayesian inference.
Bayesian phylogenetic inference often requires 
that the prior probability of the phylogeny 
according to the diversification model has to be computed millions of times. 
Therefore, biologically interesting but computationally expensive tree priors 
are often not implemented, and simpler priors are used instead. 
This is not necessarily problematic, when the data are very informative 
% 00010
\new{or when the prior is truly uninformative}, 
as this will reduce the influence of the tree prior.
However, the assumption that tree prior choice is of low \new{impact} 
must first be verified.

@@

\begin{lstlisting}[language=R]
library(ggplot2)
library(pirouette)

# Create phylogeny
phylogeny <- create_exemplary_dd_tree(
  n_taxa = 6, 
  crown_age = 10,
  extinction_rate = 0.1
)

# Use standard pirouette setup
pir_params <- create_std_pir_params()

@@

# Plot
pir_plot(pir_out)

# Save
pir_save(
  phylogeny = phylogeny,
  pir_params = pir_params,
  pir_out = pir_out,
  folder_name = "example_30"
)
\end{lstlisting}

### 2020-06-19 pirouette_supplement.tex
Accept @Giappo's suggestions, simplify example code, progress in answering response letter
Process feedback

\giovanni{
  This is the example one of the reviewer was referring to. 
  The figure that was present before (still present in the overleaf files) 
  looked actually quite different from the one on the 
  repo (\url{https://raw.githubusercontent.com/richelbilderbeek/pirouette_example_26/master/errors.png}). 
  I changed it with the one found there. In case you think this is wrong you can revert it back.
}
\richel{ 
  Well spotted! Thanks to you, the figure is correct now
}
\new{
  Since the two models are very similar to each other (the BD model can be 
  turned into a Yule model just by setting the extinction parameter to 
  zero [\cite{nee1994reconstructed}]) the median discrepancy is
  almost negligible. 
  However, with respect to the previous 
  case (subsection \ref{subsec:simplest_correct_parameterization}), 
  where a Yule tree prior has been used, 
  the distributions here exhibit a greater difference.
}

@@

\giovanni{
  I am not able to give an interpretation to these figures. 
  I would expect a trend between the difference of true and twin medians 
  and mutation rate, but I cannot see anything. The fact that sometimes true 
  is smaller and some other times twin is, makes me think that these results 
  are probably not very reliable, probably due to a not sufficient number 
  of replicates.
}
\richel{
  I interpreted our curvy friends :-)
}

>>>>

\new{
  Since the two models are very similar to each other (the BD model can be 
  turned into a Yule model just by setting the extinction parameter to 
  zero [\cite{nee1994reconstructed}]) the median discrepancy is
  almost negligible. 
  However, with respect to the previous 
  case (subsection \ref{subsec:simplest_correct_parameterization}), 
  where a Yule tree prior has been used, 
  the distributions here exhibit a greater difference.
}
\new{
  As we use only extant trees, it is reasonable that the method
  is slightly weaker in distinguishing between the Yule and BD
  models. It is unknown what the discriminatory power would be
  when comparing trees with extinction events.
}

@@
